Experiences with Planning for Natural Language Generation

### 2009-06-01 trunk/Papers/ci-spark/domains.tex
gen as planning
figures moved to the right places
sect 3

We will now focus the discussion towards two specific NLG problems:
sentence generation in the sense of \cite{KolSto07}, and the
generation of instructions in virtual environments
\cite{ByrKolStrCasDalMooObe09}.

\subsection{Sentence generation as planning}

One way of modelling the sentence generation problem is to assume a
lexicalized grammar in which each lexicon entry specifies how it can
be combined grammatically with the other lexicon entries, what piece
of meaning it expresses, and what the pragmatic conditions on using it
are. Sentence generation can then be seen as constructing a
grammatical derivation that is syntactically complete, respects the
semantic and pragmatic conditions, and achieves all the
\emph{communicative goals}. An example is the tree-adjoining grammar
(TAG; \citep{joshi;etal1997}) shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. Let's say that
we assume a knowledge base containing the individuals $e$, $r_1$ and
$r_2$ and the facts that $r_1$ and $r_2$ are rabbits, $r_1$ is white
and $r_2$ is brown, and $e$ is an event in which $r_1$ sleeps. Then we
would want to construct a sentence expressing the information
$\{\mathsf{sleep}(e,r_1)\}$ by combining instances of the elementary
trees (in which the \emph{semantic roles}, such as $\mathsf{self}$ and
$\mathsf{subj}$, have been substituted by constants from the knowledge
base) into a TAG derivation as as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv}. We can then read off the
sentence ``The white rabbit sleeps'' from the derivation. \todo{need
  to mention distractors and unique REs here}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-grammar}
  \caption{An example grammar in the sentence generation domain.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-derivation}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

This perspective on sentence generation has the advantage of solving
the sentence planning and surface realization problems simultaneously:
in the example, we can require that referring expressions ``the white
rabbit'' can be resolved uniquely to $r_1$ by the hearer, in addition
to the requirement that the derivation is grammatically correct. This
is useful in cases where sentence planning and surface realization
interact, because syntactic information about the individual words is
available when the REs are generated (see
e.g. \cite{stone98textual}). However, the problem of whether a given
communicative goal can be achieved with a given grammar is NP-complete
\cite{KolStr02}. A naive search algorithm that computes the derivation
top-down takes exponential time and is clearly infeasible to use in
practice. The SPUD system \cite{Stone2003a}, which worked this idea
out for the first time, uses a greedy algorithm, which circumvents the
combinatorial explosion but is incomplete.

\begin{figure}[p]
\centering
\begin{minipage}{0.5\textwidth}
{\small%
\begin{verbatim}
(:action add-sleeps
   :parameters (?u - node
                ?xself - individual
                ?xsubj - individual)
   :precondition
       (and (subst S ?u)
            (referent ?u ?xself)
            (sleep ?xself ?xsubj))
   :effect 
       (and (not (subst S ?u))
            (expressed sleep ?xself ?xsubj)
            (subst NP (subj ?u))
            (referent (subj ?u) ?xsubj)
            (forall (?y - individual)
                (when (not (= ?y ?xself))
                    (distractor (subj ?u) ?y)))))

@@

Here we will look at one recent approach which aims at controlling the
search by converting the sentence generation problem into a planning
problem and then running a planner \citet{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.} Each operator of this
planning problem models the addition of some elementary tree to the
TAG derivation. For instance,
Figure~\ref{fig:white-rabbit-as-planning} shows the corresponding PDDL
actions for the above generation task. The $\addact{sleeps}$ operator
represents the addition of the elementary tree for ``sleeps'' to the
derivation. $\addact{sleeps}(\mathsf{root},e,r_1)$ removes the atom
$\mathsf{subst}(S,\mathsf{root})$, which expresses that there is
currently an open substitution node $\mathsf{root}$ with label $S$,
from the planning state and replaces it with an atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$; here
$\mathsf{subj}(\mathsf{root})$ is meant as shorthand for a fresh
individual name.\footnote{These terms, which are not valid in ordinary
  PDDL, can be eliminated by estimating an upper bound $n$ for the
  plan length, making $n$ copies of each action, ensuring that copy
  $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. Notice that $S$, $NP$, and $N$ are constants.}
At the same time, the operator records that the semantic information
$\mathsf{sleep}(e,r_1)$ has now been expressed, and introduces all
individuals except for $r_1$ as distractors for the new RE at
$\mathsf{subj}(\mathsf{root})$. These distractors can then be removed
by subsequent applications of the other two operators. By assuming an
initial state that encodes the knowledge base and contains an atom
$\mathsf{subst}(S,\mathsf{root})$, and a goal that contains, among
others, $\forall x \forall y. \neg \mathsf{subst}(x,y)$ and
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$, a planner will find plans
like the following:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed. Thus, we can solve the sentence generation problem via
the detour through planning and bring current search heuristics for
planning to bear on generation.

\subsection{Planning in instruction giving}

Let's now turn to a second recent application of planning in NLG, the
GIVE Challenge (``Generating Instructions in Virtual Environments'';
\citealt{ByrKolStrCasDalMooObe09}). The object of this shared task is
to build an NLG system which produces natural-language instructions
which will guide a human user in performing some task in a virtual
environment.  From an NLG perspective, GIVE makes for an interesting
challenge because it is a theory-neutral task that exercises all
components of an NLG system, and emphasizes the study of communication
in a (simulated) physical environment. Another advantage is that GIVE
makes it possible to connect NLG systems to users over the Internet,
and thus offer cheap access to human experimental subjects. The first
installment of GIVE (GIVE-1) evaluated five NLG systems on the
performance of 1143 users, making it the largest NLG evaluation effort
in terms of human users ever.

@@

Planning plays a central role in the GIVE task. To see this, consider
the example GIVE world map in
Figure~\ref{fig:give-development-world}. To simplify both the planning
and the NLG task, this world is discretised into a set of tiles of
equal size.  The user can turn by 90 degree steps in either direction,
and can move from the centre of one tile to the centre of the next
tile, provided the path between two tiles is not blocked. The world
also consists of a set of objects which can be manipulated by the user
in various ways. For instance, in the example world the user's task is
to pick up a trophy in the top left room. The trophy is hidden in a
safe behind a picture; to access it, the user must push certain
buttons in order to move the picture out of the way, open the safe,
and open doors. Figure~\ref{fig:give-planning} shows the encoding of
some of the available GIVE domain actions, in PDDL syntax. In the
example, the shortest plan to solve the task consists of 108 action
steps, with the first few steps as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
\mathsf{west})$,
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
\end{enumerate}

\begin{figure}[p]
\centering
\begin{minipage}{0.5\textwidth}
{\small%
\begin{verbatim}
(:action move
   :parameters (?from - position
                ?to - position
                ?ori - orientation)
   :precondition 
       (and (player-pos ?from) 
            (adjacent ?from ?to ?ori) 
            (player-orient ?ori)
            (not-blocked ?from ?to)
            (not-alarmed ?to))
   :effect 
       (and (not (player-pos ?from))
            (player-pos ?to)))

@@

A GIVE NLG system must be able to compute such plans. At a minimum,
the discourse planner will call a planner in order to determine the
content of the instructions that should be presented to the
user. Under this view, the planning problem is very similar to the
Gridworld problem (see, e.g., \citep{Tovey-Koenig:2000}), which also
involves finding a route through a two-dimensional world map with
discrete positions. GIVE tasks are usually much more complex, however,
as illustrated by the above example map: a successful plan must
include steps for pressing buttons in the right order, reasoning about
large numbers of world objects, and navigating through complicated
room shapes. This relatively loose integration of NLG system and
planner is the state of the art of the systems that participated in
GIVE-1.

However, it is desirable to integrate the planner and the generation
system more closely than this. For instance, let's say that the NLG
system wants to generate the instruction sequence ``walk to the centre
of the room; turn right; now press the green button in front of you''.
Experiments \todo{cite} show that this is a pattern that human
instruction givers use frequently: The instruction follower is made to
walk to a certain point in the world at which the instruction giver
can then use a a referring expression (``the green button'') that is
very easy to interpret. Such an NLG system must integrate discourse
planning and planning in the domain of the world map closely: On the
one hand, the structure of the discourse is determined by the needs of
the NLG system rather than the domain plan; on the other hand, the
discourse planner must be aware of the way in which the instruction
``turn right'' is likely to change the visibility of objects. Even if
the NLG system doesn't implement the generation of this discourse as
planning, it still solves a problem that subsumes the domain planning
problem. For these reasons, we consider the GIVE domain planning
problem as a natural part of a GIVE NLG system.
\todo{Make sure that this paragraph still talks about the things that
  the reviewers found interesting about it.}

>>>>

We will now focus the discussion on two specific NLG problems:
sentence generation in the sense of \citet{KolSto07}, and the
generation of instructions in virtual environments
\citep{ByrKolStrCasDalMooObe09}. For each, we will introduce the task
and show by example how the task can be seen as a planning problem.

\subsection{Sentence generation as planning}
\label{sec:domain-crisp}

One way of modelling the sentence generation problem is to assume a
lexicalized grammar in which each lexicon entry specifies how it can
be combined grammatically with the other lexicon entries, what piece
of meaning it expresses, and what the pragmatic conditions on using it
are. Sentence generation can then be seen as constructing a
grammatical derivation that is syntactically complete, respects the
semantic and pragmatic conditions, and achieves all the
\emph{communicative goals}.

An example is the tree-adjoining grammar (TAG;
\citealt{joshi;etal1997}) shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. Let's say that
we assume a knowledge base containing the individuals $e$, $r_1$ and
$r_2$ and the facts that $r_1$ and $r_2$ are rabbits, $r_1$ is white
and $r_2$ is brown, and $e$ is an event in which $r_1$ sleeps. Then we
would want to construct a sentence expressing the information
$\{\mathsf{sleep}(e,r_1)\}$ by combining instances of the elementary
trees (in which the \emph{semantic roles}, such as $\mathsf{self}$ and
$\mathsf{subj}$, have been substituted by constants from the knowledge
base) into a TAG derivation as as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv}. In the picture, the dashed
arrow indicates TAG's \emph{substitution} operation, which ``plugs''
an elementary into the leaf of another tree, and the dotted arrows
stand for \emph{adjunction}, which splices an elementary tree into an
internal node. We can then read off the sentence ``The white rabbit
sleeps'' from the derivation. Note that the sentence ``The rabbit
sleeps'' would not have been an appropriate result, because ``the
rabbit'' could refer to either $r_1$ or $r_2$, and thus $r_2$ remains
as a \emph{distractor}, i.e.\ an incorrect possible interpretation of
the phrase.

\begin{figure}
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-grammar}
  \caption{An example grammar in the sentence generation domain.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-derivation}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

This perspective on sentence generation has the advantage of solving
the sentence planning and surface realization problems
simultaneously. The generation of referring expressions (REs) is
usually seen as a sentence planning task; in the example, we required
that the referring expression ``the white rabbit'' can be resolved
uniquely to $r_1$ by the hearer, in addition to the requirement that
the derivation is grammatically correct. This is useful in cases where
sentence planning and surface realization interact, because syntactic
information about the individual words is available when the REs are
generated (see e.g. \citealt{stone98textual}). However, the problem of
whether a given communicative goal can be achieved with a given
grammar is NP-complete \citep{KolStr02}. A naive search algorithm that
computes the derivation top-down takes exponential time and is clearly
infeasible to use in practice. The seminal SPUD system
\citep{Stone2003a}, which first worked out the idea of integrated
TAG-based sentence generation, used a greedy algorithm to cope with
the search, which circumvents the combinatorial explosion but is
incomplete.

\begin{figure}
\centering
\begin{minipage}{0.8\textwidth}
{\small%
\begin{verbatim}
(:action add-sleeps
   :parameters (?u - node  ?xself - individual  ?xsubj - individual)
   :precondition
       (and (subst S ?u)  (referent ?u ?xself)  (sleep ?xself ?xsubj))
   :effect 
       (and (not (subst S ?u))  (expressed sleep ?xself ?xsubj)
            (subst NP (subj ?u))  (referent (subj ?u) ?xsubj)
            (forall (?y - individual)
                (when (not (= ?y ?xself)) (distractor (subj ?u) ?y)))))

@@

As an alternative, \citet{KolSto07} recently proposed to control the
search by converting the sentence generation problem into a planning
problem and then running a planner \citet{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.} This planning problem assumes
an initial state containing an atom $\mathsf{subst}(S,\mathsf{root})$,
encoding that we must generate a sentence ($S$) starting at the node
named $\mathsf{root}$ in the TAG derivation tree, and an atom
$\mathsf{referent}(\mathsf{root},e)$ encoding that the (event)
individual which the entire sentence describes is $e$. Then it encodes
the addition of elementary trees to the TAG derivation in the
individual planning operators; the ones that are needed for our
example above are shown in
Fig.~\ref{fig:white-rabbit-as-planning}.

Here the action instance $\addact{sleeps}(\mathsf{root}, e, r_1)$
replaces the atom $\mathsf{subst}(S,\mathsf{root})$ with the atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
syntax, we write $\mathsf{subj}(\mathsf{root})$ as shorthand for a
fresh individual name.\footnote{These terms, which are not valid in
  ordinary PDDL, can be eliminated by estimating an upper bound $n$
  for the plan length, making $n$ copies of each action, ensuring that
  copy $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ in the planning
  problem are constants.}  At the same time, the operator records that
the semantic information $\mathsf{sleep}(e,r_1)$ has now been
expressed, and introduces all individuals except for $r_1$ as
distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
distractors can then be removed by subsequent applications of the
other two operators. Eventually we reach a goal state, which is
characterized by goals including $\forall x \forall y. \neg
\mathsf{subst}(x,y)$, $\forall x \forall y. \neg
\mathsf{distractor}(x,y)$, and
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. In summary, the following
is a correct plan:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
The grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed from this plan. Thus, we can solve the sentence
generation problem via the detour through planning and bring current
search heuristics for planning to bear on generation.

\subsection{Planning in instruction giving}
\label{sec:domain-give}

Let us now turn to a second recent application of planning in NLG, the
GIVE Challenge (``Generating Instructions in Virtual Environments'';
\citealt{ByrKolStrCasDalMooObe09}). The object of this shared task is
to build an NLG system which produces natural-language instructions
which will guide a human user in performing some task in a virtual
environment.  From an NLG perspective, GIVE makes for an interesting
challenge because it is a theory-neutral task that exercises all
components of an NLG system, and emphasizes the study of communication
in a (simulated) physical environment. Furthermore, because the client
which displays the 3D environment to the user and the NLG system can
be physically separated, as long as they are connected over a network,
such systems can be cheaply evaluated over the Internet. This provides
a potential solution to the long-standing problem of evaluating NLG
systems. The first installment of GIVE (GIVE-1) evaluated five NLG
systems on the performance of 1143 users, making it the largest NLG
evaluation effort in terms of human users ever.

@@

Planning plays a central role in the GIVE task. To see this, consider
the example GIVE world map in
Figure~\ref{fig:give-development-world}. In this example world, the
user's task is to pick up a trophy in the top left room. The trophy is
hidden in a safe behind a picture; to access it, the user must push
certain buttons in order to move the picture out of the way, open the
safe, and open doors. The user must to navigate the world and perform
these actions in the 3D client in a certain order; the NLG system must
instruct the user on how to do this.  To simplify both the planning
and the NLG task, the world is discretised into a set of tiles of
equal size. The user can turn by 90 degree steps in either direction,
and can move from the centre of one tile to the centre of the next
tile, provided the path between two tiles is not blocked.
Figure~\ref{fig:give-planning} shows the encoding of some of the
available GIVE domain actions, in PDDL syntax. In the example, the
shortest plan to solve the task consists of 108 action steps, with the
first few steps as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
\mathsf{west})$,
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
\end{enumerate}

\begin{figure}
\centering
\begin{minipage}{0.5\textwidth}
{\small%
\begin{verbatim}
(:action move
   :parameters (?from - position
                ?to - position
                ?ori - orientation)
   :precondition 
       (and (player-pos ?from) 
            (adjacent ?from ?to ?ori) 
            (player-orient ?ori)
            (not-blocked ?from ?to)
            (not-alarmed ?to))
   :effect 
       (and (not (player-pos ?from))
            (player-pos ?to)))

@@

A GIVE NLG system must be able to compute such plans. At a minimum,
the discourse planner will call a planner in order to determine the
content of the instructions that should be presented to the
user. Under this view, the planning problem is very similar to the
Gridworld problem (see, e.g., \citealt{Tovey-Koenig:2000}), which also
involves finding a route through a two-dimensional world map with
discrete positions. As in Gridworld, the domain requires not only
route finding in the discretised world, but also the execution of
certain actions (such as finding keys and unlocking doors in
Gridworld, or pushing the correct buttons to open doors and the safe
in GIVE). However, the GIVE worlds tend to be bigger than the
Gridworld instances that were used in the 1990s planning competitions.

This relatively loose integration of NLG system and planner is the
state of the art of the systems that participated in GIVE-1. However,
it is desirable to integrate the planner and the generation system
more closely than this. For instance, let's say that the NLG system
wants to generate the instruction sequence ``walk to the centre of the
room; turn right; now press the green button in front of
you''. Experiments with human instruction givers \cite{stoia:08} show
that this is a pattern that they use frequently: The instruction
follower is made to walk to a certain point in the world at which the
instruction giver can then use a a referring expression (``the green
button'') that is very easy to interpret. Such an NLG system must
integrate discourse planning and planning in the domain of the world
map closely. On the one hand, the structure of the discourse is
determined by the needs of the NLG system rather than the domain plan;
on the other hand, the discourse planner must be aware of the way in
which the instruction ``turn right'' is likely to change the
visibility of objects. Even if the NLG system doesn't implement the
generation of this discourse as planning, it still solves a problem
that subsumes the domain planning problem. For these reasons, we
consider the GIVE domain planning problem as a natural part of a GIVE
NLG system.  \todo{Make sure that this paragraph still talks about the
  things that the reviewers found interesting about it.}

### 2009-06-01 trunk/Papers/ci-spark/introduction.tex
sect 3

The remainder of this paper is structured as follows. In
Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as planning and
briefly review the relevant literature. In Section~\ref{sec:domains}, we
describe a set of planning problems associated with two NLG tasks, namely
sentence planning and instruction generation. In Section~\ref{sec:experiments},
we report on the results of our experiments with these planning problems.
In Section~\ref{sec:discussion} we discuss our results and overall
experiences, and conclude in Section~\ref{sec:conclusion}.

>>>>

The remainder of this paper is structured as follows. In
Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as
planning and briefly review the relevant literature. In
Section~\ref{sec:domains}, we describe a set of planning problems
associated with two NLG tasks, namely sentence planning and situated
instruction generation. In Section~\ref{sec:experiments}, we report on
the results of our experiments with these planning problems.  In
Section~\ref{sec:discussion} we discuss our results and overall
experiences, and conclude in Section~\ref{sec:conclusion}.
